---
title: "Introduction to Poisson processes with R"
author: "5 ModIA"
date: "2023/2024"
output: 
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


\section{1 - Homogeneous Poisson processes observed on a fixed window}

First, we consider the case of a fixed observation window (and thus a random number of events). 

\subsection{1.1 - Simulation}

```{r,eval=FALSE}
simulPPh1 <- function(lambda,Tmax)
{
  Poisson = rpois(1, lambda=lambda*Tmax)
  unif = runif(Poisson, min = 0, max=Tmax)
  unif_trie = sort(unif)
  return(unif_trie)
}
```


```{r,eval=FALSE}
# simulate a homogeneous Poisson process:
PPh1 = simulPPh1(lambda = 2, Tmax = 10)

# plot the counting process (with jumps = 1): 
plot(c(0,PPh1), 0:length(PPh1),type="s",xlab="time t",ylab="number of events by time t")

# add the arrival times on the horizontal axis: 
points(PPh1, 0*PPh1,type="p",pch=16)

# link the arrival times with the counts:
lines(PPh1,1:length(PPh1),type="h",lty=2)
```


\subsection{1.2 - Maximum Likelihood Estimator (MLE)}

```{r,eval=FALSE}
MLE1 <- function(PPh,Tmax)
{ NTmax = length(PPh)
  return(NTmax/Tmax)
}
```

```{r,eval=FALSE}
MLE1(simulPPh1(lambda = 2, Tmax = 10), Tmax = Tmax)
```

Comment:

\subsection{1.3 - Asymptotic behavior of the MLE}

\subsubsection{1.3.1 - LLN-type result}

```{r,eval=FALSE}
lambda = 2
Tillustr =1:500
mle_values <- list()
pph2_list <- list()
for(Tmax in Tillustr){
  Pph2 = simulPPh1(lambda,Tmax)
  pph2_list = append(pph2_list, simulPPh1(lambda,Tmax))
  mle_values = append(mle_values, MLE1(Pph2, Tmax))
}

plot(Tillustr, mle_values)
```

Comment: 


\subsubsection{1.3.2 - CLT-type result}

```{r,eval=FALSE}
"COMPLETE"
```

Comment: 


\subsection{1.4 - Statistical inference: hyptothesis testing}

```{r,eval=FALSE}
test1 <- function(PPh,Tmax,lambda0)
{
  "COMPLETE"
  return("COMPLETE")
}
```

```{r}
# Plot confidence intervals for the proportion of times alpha-level tests rejects the 
# null hypothesis "lambda=lambda0" under each true distribution lambda in TrueLambda
plot.level.power1 <- function(Tmax,lambda0,TrueLambda,alpha,nsimu)
{
  plot(range(TrueLambda),c(alpha,alpha),ylim=c(0,1),xlab="True lambda",ylab="Level/Power",
       type="l",col="red",main=paste("lambda0 = ",lambda0,",  Tmax = ",Tmax,sep=""))
  abline(1,0,lty=2,col="blue")
  
  for(lambda in TrueLambda)
  {
    # estimating the proportion under lambda in TrueLambda
    propReject=0
    for(sim in 1:nsimu){
      propReject=propReject+(test1(simulPPh1(lambda,Tmax),Tmax,lambda0) <= alpha)/nsimu
    }
  # plot the confidence intervals
  points(lambda,propReject)
  points(lambda,propReject+sqrt(abs(propReject*(1-propReject))/nsimu)*qnorm(0.975),pch=2)
  points(lambda,propReject-sqrt(abs(propReject*(1-propReject))/nsimu)*qnorm(0.975),pch=6)  
  }
}
```

```{r,eval=FALSE}
alpha="COMPLETE"
nsimu="COMPLETE"
lambda0="COMPLETE"
TrueLambda=c("COMPLETE")

par(mfrow=c(1,2))
for(Tmax in c(1,10,100,500))
{
  plot.level.power1("COMPLETE")
}
```

Comments:

\section{2 - Homogeneous Poisson processes with fixed number of points}

Second, we consider the case of a fixed number of points (and thus a random observation window). 

\subsection{2.1 - Simulation}

```{r,eval=FALSE}
simulPPh2 <- function(lambda,n)
{
  return("COMPLETE")
}
```

```{r,eval=FALSE}
PPh2 = "COMPLETE"
"COMPLETE" 
```

\subsection{2.2 - Maximum likelihood estimator}

```{r,eval=FALSE}
MLE2 <- function(PPh)
{
  return("COMPLETE")
}

MLE2("COMPLETE")
```

Comment:

\subsection{2.3 Asymptotic behavior of the MLE}

\subsubsection{2.3.1 - LLN-type result}

```{r,eval=FALSE}
"COMPLETE"
```

Comment: 

\subsubsection{2.3.2 - CLT-type result}

```{r,eval=FALSE}
"COMPLETE"
```

Comment:

\subsection{2.4 - Statistical inference : confidence intervals}

```{r,eval=FALSE}
IC2 <- function(PPh,alpha=0.05,asymptotic)
{
  if(asymptotic)
  {
    "COMPLETE"
    return("COMPLETE")
  }
  else
  {
    "COMPLETE"
    return("COMPLETE")
  }
}
```

```{r,eval=FALSE}
# Application on an example
PPh2 <- simulPPh2(lambda=10,n=100)
IC2(PPh2,alpha=0.05,asymptotic=TRUE)
IC2(PPh2,alpha=0.05,asymptotic=FALSE)
```

Comment:

```{r,eval=FALSE}
# Validation on simulated data
lambda=2 ; nsimu=1000 ; n=10 # or n=100
"COMPLETE"
```

Comment:

\section{3 - Inhomogeneous Poisson processes}

Third, we simulate inhomogeneous Poisson processes with given intensity function on a fixed window. 


```{r,eval=FALSE}
simulPPi = function(lambda_fct,Tmax,M)
{
  n = rpois(1, M*Tmax)
  u = runif(n, 0, Tmax)
  y = runif(n,0,M)
  t = u[y<=lambda_fct(u)]
  return(sort(t))
}
```

```{r,eval=FALSE}
Tmax= 10
lambda_fct1 <- function(t){return("COMPLETE")}
M1="COMPLETE"

PPi1 = simulPPi("COMPLETE")
"COMPLETE"
```

```{r,eval=FALSE}
Tmax="COMPLETE"
lambda_fct2 <- function(t){return("COMPLETE")}
M2="COMPLETE"

PPi2 = simulPPi("COMPLETE")
"COMPLETE"
```




